Wired for Success: Sweden's Electrical Innovators in the Age of Industrial Transformation

### 2023-09-29 index.qmd
add umap

---
title: "Wired for Success: Sweden's Electrical Innovators in the Age of Industrial Transformation"
subtitle: "CBS Department of Strategy and Innovation Graduate Seminar"
author: "Jonathan Jayes  - Lund University"
title-slide-attributes:
  data-background-image: "assets/preview.png"
  data-background-size: "30%"
  data-background-opacity: "0.6"
format:

  revealjs:
    code-overflow: wrap
    width: 1600
    height: 900
    fig-align: center
    default-image-extension: svg
    footer: "[{{< fa house >}}](https://j-jayes.github.io/Data-Visualization-in-R-Lund/)"
    slide-number: c
    code-block-bg: true
    code-block-border-left: "#31BAE9"
image: "preview_img/03-coordinate-systems-and-axes.jfif"
favicon: html/favicon.ico
---

@@

    - Who were the innovators or adopters? 🤔🚀
    - Did they move to opportunity? 🏃‍♀️💼
    - How much training did they have in the new technology? 🎓⚡️

@@

3. It is a General Purpose Technology that can both process data and also inform our workflow as researchers.

@@

## Why do we care about the second industrial revolution?

![](assets/factory.png){fig-align="center"}

![](assets/2IR_1.JPG){fig-align="center"}

![](assets/2IR_2.JPG){fig-align="center"}

## New technology with rapid adoption in Sweden

![](assets/schon-energy-share.jpeg){fig-align="center"}

## What happened to early electrifiers in Sweden?

![](assets/grid_rollout.jpg){fig-align="center"}

## No large unemployment effects in areas who electrified first

![](assets/employment_share.jpeg){fig-align="center"}

## Incomes were markedly higher in areas that electrified first

![](assets/coef_plot.jpeg){fig-align="center"}

## Income was shared more equally in areas that electrified first

![](assets/dot_plot.jpeg){fig-align="center"}

## What were the mechanisms behind these effects?

![](assets/job_share_plot.jpeg){fig-align="center"}

## Now we want to ask:

- Who were the innovators or adopters? 🤔🚀
- Did they move to opportunity? 🏃‍♀️💼
- How much training did they have in the new technology? 🎓⚡️

@@

- Biographies and career trajectories of ~ 69,448 individuals!

- 8 of the 10 volumes are digitized by librarians in Uppsala – thank you <3

@@

# Create a tibble from the JSON data
books <- tibble(
  book_name = c("Skåne", "Stockholm", "Svealand", "The Götaland region except Skåne", "Skåne", "Norrland", "Stockholm", "The Götaland region except Skåne"),
  year = c(1948, 1962, 1964, 1965, 1966, 1968, 1945, 1948),
  pages = c(624, 1498, 945, 1203, 952, 1340, 1044, 1075),
  volume = c(1, 2, 2, 2, 2, 2, 1, 1)
)

books <- books %>%
    mutate( name = str_c(book_name, " - ", year),
            name = fct_reorder(name, pages),
            records = pages * 8) %>%
            mutate(volume = if_else(volume == 1, "1945–1950", "1962–1968")) %>%
    select(name, records, volume)

@@

# Create the ggplot visualization
ggplot(books, aes(x = records, y = name, fill = factor(volume))) +
  geom_col(alpha = .8) +
  labs(x = "Number of biographies", y = "Book volume",
  title = "Number of biographies in Vem är Vem?",
  fill = "Edition") +
  scale_x_continuous(labels = scales::number_format()) +
  scale_fill_manual(values = c("#2B8CBE", "#be5d2b"))

@@

Lund, Karl Gustaf, chief engineer, Varberg, born on July 22, 1893, in Hille, Gävle County, Sweden, son of factory worker Ferdinand L. and Maria Andersson. Married in 1936 to Sigrid Johansson. Children: Ingvar (born 1938), Lennart (born 1942). — Graduated from Bergsskolan in Filipstad in 1917, specialized studies at the Royal Institute of Technology (KTH) from 1920 to 1922, studied at the Institute of Metallurgy and Stockholm University in 1921-1922. Chemist at Strömsnäs Järnverks A-B in Degerfors from 1918 to 1920, metallurgist and chemist at Westinghouse Electric & Manufacturing Co. in East Pittsburgh, PA, USA, from 1923 to 1926 and 1928 to 1929, chief metallurgist at Laclede Steel Co. in Alton, Illinois, USA, in 1927, furnace and steelworks engineer at A-B Iggesunds Bruk from 1929 to 1931, site manager at Gunnebo Bruks Nya A-B, Varbergsverket, since 1931. Member of the municipal executive committee, deputy chairman of the economic department, deputy member of the board of the power plant, chairman of Varbergs Sparbank, employer representative in the district council of the county labor board, member of the board of Varbergs Luftskyddsförening (Varberg Air Protection Association), secretary of Varbergs Högerförening (Varberg Conservative Association), chairman of the railway sick fund, and Plant Society for Small Bird Friends.

Traveled to Germany in 1921, 1923, 1930, and 1936, Denmark, Czechoslovakia in 1921, 1922, and 1923, Austria in 1921, and the USA from 1923 to 1929. Writings: "Some fundamental factors for obtaining sharp thermal curves" (Transactions of the American Society for Steel Treating, co-authored with C. Benedicks and W. H. Dearden, 1925), "Contemporary production of saw blades, circular saws, and machine knives" (Timber Industry, 1931). Hobbies: hunting and fishing.

[Lund, Karl Gustaf]{style="background-color: #b0c4de"}, [chief engineer]{style="background-color: #d2b48c"}, [born on July 22, 1893]{style="background-color: #ff8c69"}, [in Hille, Gävle County]{style="background-color: #90ee90"}, Sweden, son of [factory worker Ferdinand L. and Maria Andersson]{style="background-color: #9370db"}. Married in 1936 to [Sigrid Johansson]{style="background-color: #ffadc1"}. Children: [Ingvar (born 1938)]{style="background-color: #ffdead"}, [Lennart (born 1942)]{style="background-color: #ffdead"}. — Graduated from [Bergsskolan]{style="background-color: #afeeee"} in Filipstad in 1917, specialized studies at the [Royal Institute of Technology (KTH)]{style="background-color: #afeeee"} from 1920 to 1922, studied at the [Institute of Metallurgy]{style="background-color: #afeeee"} and [Stockholm University]{style="background-color: #afeeee"} in 1921-1922. Chemist at [Strömsnäs Järnverks A-B]{style="background-color: #98fb98"} in Degerfors from 1918 to 1920, metallurgist and chemist at [Westinghouse Electric & Manufacturing Co.]{style="background-color: #98fb98"} in East Pittsburgh, PA, USA, from 1923 to 1926 and 1928 to 1929, chief metallurgist at [Laclede Steel Co.]{style="background-color: #98fb98"} in Alton, Illinois, USA, in 1927, furnace and steelworks engineer at [A-B Iggesunds Bruk]{style="background-color: #98fb98"} from 1929 to 1931, site manager at [Gunnebo Bruks Nya A-B, Varbergsverket]{style="background-color: #98fb98"}, since 1931. Member of the municipal executive committee, deputy chairman of the economic department, deputy member of the board of the power plant, chairman of [Varbergs Sparbank]{style="background-color: #ffdab9"}, employer representative in the district council of the county labor board, member of the board of [Varbergs Luftskyddsförening (Varberg Air Protection Association)]{style="background-color: #ffdab9"}, secretary of [Varbergs Högerförening (Varberg Conservative Association)]{style="background-color: #ffdab9"}, chairman of the railway sick fund, and [Plant Society for Small Bird Friends]{style="background-color: #ffdab9"}.

@@

```{python, eval=FALSE, echo=TRUE}
#| code-line-numbers: "6-14"

@@

    "translated": "Lund, Karl Gustaf, chief engineer, Varberg, born on July 22, 1893, in Hille, Gävle County, Sweden, son of factory worker Ferdinand L. and Maria Andersson. Married in 1936 to Sigrid Johansson. Children: Ingvar (born 1938), Lennart (born 1942). — Graduated from Bergsskolan in Filipstad in 1917, specialized studies at the Royal Institute of Technology (KTH) from 1920 to 1922, studied at the Institute of Metallurgy and Stockholm University in 1921-1922. Chemist at Strömsnäs Järnverks A-B in Degerfors from 1918 to 1920.",

@@

        "@context": "https://schema.org",

@@

@@

            "name": "Hille, Gävle County, Sweden",
            "latitude": 59.916667,
            "longitude": 15.0

@@

        "jobTitle": "Chief Engineer",
        "memberOf": {
            "name": "Swedish Technical Association"

@@

                "name": "Anita",
                "birthDate": "1937"

@@

                "name": "Peter",

@@

            "name": "Marianne Hammarberg",
        "education": [

@@

                "@type": "EducationalOccupationalCredential",
                "credentialCategory": "Vocational",
                "issuer": {
                    "name": "Technical Gymnasium in Örebro"
                "endDate": "1928"

@@

                "@type": "EducationalOccupationalCredential",
                "credentialCategory": "Degree",
                "issuer": {
                    "name": "KTH"

@@

                "endDate": "1934"

@@

        "worksFor": {

@@

            "name": "Telegraph station in Norrköping",
            "startDate": "1946"
        "hasOccupation": [
                "name": "Engineering assistant",
                "employer": {
                    "name": "Telegraph station in Nässjö"
                "startDate": "1935",
                "endDate": "1935"

@@

@@

                "name": "Line engineer",
                "employer": {
                    "name": "Telegraph station in Norrköping"

@@

                "startDate": "1946"

@@

        ]

@@

## Questions I need to answer

1. What is an innovator in this context?

2.

## How were NLP tasks hard before?

In your case when you use patent data, a lot of the research on historical questions is done just with the title of the patent and relationships between patent holders. It was difficult to get more information from the text of the patent.

### How are they easy now?

Now we can use OCR / NLP to get a lot of information from the text of the patent.

>>>>

---
title: "Wired for Success: Sweden's Electrical Innovators in the Age of Industrial Transformation"
subtitle: "SEHM 2023: The application of cutting edge technology to economic history research"
author: "Jonathan Jayes  - Lund University"
title-slide-attributes:
  data-background-image: "assets/preview.png"
  data-background-size: "30%"
  data-background-opacity: "0.6"
format: 
  revealjs:
    code-overflow: wrap
    width: 1600
    height: 900
    fig-align: center
    default-image-extension: svg
    footer: "[{{< fa house >}}](https://j-jayes.github.io/Data-Visualization-in-R-Lund/)"
    slide-number: c
    code-block-bg: true
    code-block-border-left: "#31BAE9"
image: "preview_img/03-coordinate-systems-and-axes.jfif"
favicon: html/favicon.ico
---

@@

    - Who were the innovators or adopters? 🤔🚀
    - Did they move to opportunity? 🏃‍♀️💼
    - How much training did they have in the new technology? 🎓⚡️
    - Were did they get this training? 🏫🔌

# Research question{background-color="#1C9E77"}

How important was **migration and skill adoption** for the **electricity adjascent technology development** in Sweden?

@@

3. It is a **General Purpose Technology** that can both process data and also inform our workflow as researchers.

@@

@@

- Biographies and career trajectories of ~ 75,000 individuals!

- 9 of the 11 volumes are digitized by librarians in Uppsala – thank you <3

@@

# Create a tibble from the JSON data
books <- tibble(
  book_name = c("Skåne", "Stockholm", "Svealand", "The Götaland region except Skåne", "Skåne", "Norrland", "Stockholm", "The Götaland region except Skåne", "Who is who in indsutry and business?"),
  year = c(1948, 1962, 1964, 1965, 1966, 1968, 1945, 1948, 1945),
  pages = c(624, 1498, 945, 1203, 952, 1340, 1044, 1075, 625),
  volume = c(1, 2, 2, 2, 2, 2, 1, 1, 3)
)

books <- books %>%
    mutate(name = str_c(book_name, " - ", year),
            name = fct_reorder(name, pages),
            records = pages * 8) %>%
            mutate(volume = case_when(
                volume == 1 ~ "1945–1950", 
                volume == 2 ~"1962–1968",
                TRUE ~ "Who is who in indsutry and business?")
            ) %>%
    select(name, records, volume)

@@

# Create the ggplot visualization
ggplot(books, aes(x = records, y = name, fill = factor(volume))) +
  geom_col(alpha = .8) +
  labs(x = "Number of biographies", y = "Book volume",
  title = "Number of biographies in Vem är Vem?",
  fill = "Edition") +
  scale_x_continuous(labels = scales::number_format()) +
  scale_fill_manual(values = c("#2B8CBE", "#be5d2b", "midnightblue"))

@@

[Lund, Karl Gustaf]{style="background-color: #b0c4de"}, [chief engineer]{style="background-color: #d2b48c"}, [born on July 22, 1893]{style="background-color: #ff8c69"}, [in Hille, Gävleborgs län]{style="background-color: #90ee90"}, Sweden, son of [factory worker Ferdinand L. and Maria Andersson]{style="background-color: #9370db"}. Married in 1936 to [Sigrid Johansson]{style="background-color: #ffadc1"}. Children: [Ingvar (born 1938)]{style="background-color: #ffdead"}, [Lennart (born 1942)]{style="background-color: #ffdead"}. — Graduated from [Bergsskolan]{style="background-color: #afeeee"} in Filipstad in 1917, specialized studies at the [Royal Institute of Technology (KTH)]{style="background-color: #afeeee"} from 1920 to 1922, studied at the [Institute of Metallurgy]{style="background-color: #afeeee"} and [Stockholm University]{style="background-color: #afeeee"} in 1921-1922. Chemist at [Strömsnäs Järnverks A-B]{style="background-color: #98fb98"} in Degerfors from 1918 to 1920, metallurgist and chemist at [Westinghouse Electric & Manufacturing Co.]{style="background-color: #98fb98"} in East Pittsburgh, PA, USA, from 1923 to 1926 and 1928 to 1929, chief metallurgist at [Laclede Steel Co.]{style="background-color: #98fb98"} in Alton, Illinois, USA, in 1927, furnace and steelworks engineer at [A-B Iggesunds Bruk]{style="background-color: #98fb98"} from 1929 to 1931, site manager at [Gunnebo Bruks Nya A-B, Varbergsverket]{style="background-color: #98fb98"}, since 1931. Member of the municipal executive committee, deputy chairman of the economic department, deputy member of the board of the power plant, chairman of [Varbergs Sparbank]{style="background-color: #ffdab9"}, employer representative in the district council of the county labor board, member of the board of [Varbergs Luftskyddsförening (Varberg Air Protection Association)]{style="background-color: #ffdab9"}, secretary of [Varbergs Högerförening (Varberg Conservative Association)]{style="background-color: #ffdab9"}, chairman of the railway sick fund, and [Plant Society for Small Bird Friends]{style="background-color: #ffdab9"}.

@@

@@

    "translated": "Lund, Karl Gustaf, chief engineer, Varberg, born on July 22, 1893, in Hille, Gävleborgs län, Sweden, son of factory worker Ferdinand L. and Maria Andersson. Married in 1936 to Sigrid Johansson. Children: Ingvar (born 1938), Lennart (born 1942). — Graduated from Bergsskolan in Filipstad in 1917, specialized studies at the Royal Institute of Technology (KTH) from 1920 to 1922, studied at the Institute of Metallurgy and Stockholm University in 1921-1922. Chemist at Strömsnäs Järnverks A-B in Degerfors from 1918 to 1920.",

@@

            "@context": "http://schema.org",

@@

            "givenName": "Karl Gustaf",
            "familyName": "Lund",
            "gender": "Male",

@@

                "@type": "Place",
                "address": {
                "@type": "PostalAddress",
                "addressLocality": "Hille",
                "addressRegion": "Gävleborgs län",
                "addressCountry": "Sweden"

@@

            "nationality": "Swedish",
            "parent": [
                "name": "Ferdinand L.",
                "jobTitle": "Factory Worker"
                "name": "Maria Andersson"
                "name": "Sigrid Johansson",

@@

                "name": "Ingvar",
                "birthDate": "1938"

@@

                "name": "Lennart",

@@

            "alumniOf": [

@@

                "@type": "EducationalOrganization",
                "name": "Bergsskolan in Filipstad",
                "graduationDate": "1917"

@@

                "@type": "EducationalOrganization",
                "name": "Royal Institute of Technology (KTH)",
                "attendancePeriod": "1920-1922"

@@

                "@type": "EducationalOrganization",
                "name": "Institute of Metallurgy and Stockholm University",
                "attendancePeriod": "1921-1922"

@@

            "worksFor": [

@@

                "name": "Strömsnäs Järnverks A-B",
                "jobTitle": "Chemist",
                "employmentPeriod": "1918-1920"

@@

                "name": "Westinghouse Electric & Manufacturing Co.",
                "jobTitle": "Metallurgist and Chemist",
                "employmentPeriod": "1923-1926, 1928-1929",
                "location": "East Pittsburgh, PA, USA"
                "name": "Laclede Steel Co.",
                "jobTitle": "Chief Metallurgist",
                "employmentPeriod": "1927",
                "location": "Alton, Illinois, USA"
                "name": "A-B Iggesunds Bruk",
                "jobTitle": "Furnace and Steelworks Engineer",
                "employmentPeriod": "1929-1931"
                "name": "Gunnebo Bruks Nya A-B, Varbergsverket",
                "jobTitle": "Site Manager",
                "employmentPeriod": "1931-present"
            "memberOf": [
                "name": "Municipal Executive Committee"
                "name": "Varbergs Sparbank",
                "position": "Chairman"
                "name": "Varbergs Luftskyddsförening",
                "position": "Member of the Board"
                "name": "Varbergs Högerförening",
                "position": "Secretary"
            "hasOccupation": {

@@

                "name": "Chief Engineer"

@@

            "address": {
                "@type": "PostalAddress",
                "addressLocality": "Varberg",
                "addressCountry": "Sweden"
            "author": [
                "@type": "CreativeWork",
                "name": "Some fundamental factors for obtaining sharp thermal curves",
                "publisher": "Transactions of the American Society for Steel Treating",
                "author": ["Karl Gustaf Lund", "C. Benedicks", "W. H. Dearden"],
                "datePublished": "1925"
                "@type": "CreativeWork",
                "name": "Contemporary production of saw blades, circular saws, and machine knives",
                "publisher": "Timber Industry",
                "author": "Karl Gustaf Lund",
                "datePublished": "1931"

@@

            "interests": "Hunting and Fishing"

@@

## How to measure career progression

```{python, eval=FALSE, echo=TRUE}
#| code-line-numbers: "10-14"
{
  "career": [
    {
      "occupational_title": "Chemist",
      "workplace": "Strömsnäs Järnverks A-B",
      "industry": "Metallurgy",
      "start_date": "1918"
    },
    {
      "occupational_title": "Metallurgist and Chemist",
      "workplace": "Westinghouse Electric & Manufacturing Co.",
      "industry": "Manufacturing",
      "start_date": "1923"
    },
    {
      "occupational_title": "Chief Metallurgist",
      "workplace": "Laclede Steel Co.",
      "industry": "Steel Production",
      "start_date": "1927"
    },
    {
      "occupational_title": "Furnace and Steelworks Engineer",
      "workplace": "A-B Iggesunds Bruk",
      "industry": "Metallurgy",
      "start_date": "1929"
    },
    {
      "occupational_title": "Site Manager",
      "workplace": "Gunnebo Bruks Nya A-B, Varbergsverket",
      "industry": "Metallurgy",
      "start_date": "1931"
    }
  ]
}

## Assign HISCOs: process

```{mermaid}
graph TB
    A[Collect HISCO Codes<br>and Descriptions] -->|Use OpenAI API| B[Convert HISCO Titles<br>and Descriptions to Vectors]
    C[Receive Occupational<br>Strings] -->|Use OpenAI API| D[Convert Occupational<br>Strings to Vectors]
    B --> E[Compare Vectors in<br>Vector Space]
    D --> E
    E --> F[Assign HISCO Code to<br>Occupational String<br>using Cosine Distance]
    style A fill:#2B8CBE
    style B fill:#be5d2b
    style C fill:#2B8CBE
    style D fill:#be5d2b
    style E fill:#6f6fbf
    style F fill:#F4FA58

## What is an embedding?

![](assets/word2vec-king-queen-vectors.png){fig-align="center" width=100%}

## What is an embedding?{.unnumbered}

![](assets/word2vec-king-queen-composition.png){fig-align="center" width=100%}

## Assign HISCOs: example

### 026 Metallurgists

Workers in this unit group advise on metallurgical problems and develop and control processes for the extraction of metals from their ores, study the properties and engineering characteristics of metals and develop new alloys, and develop and supervise metal manufacturing processes for making finished metal products.

### Karl Lund in 1923

Metallurgist and Chemist at Westinghouse Electric & Manufacturing Co.

## Assign HISCOs: codes projected from 1536 dimensions to 2 with UMAP

```{r}
knitr::include_url("assets/interactive_umap_plot.html", height = "800")
```

## Assign HISCOs: Watchmaker Apprentice

```{r}
library(gt)

# Create a tibble with the provided data
watchmaker_data <- tibble(
  Rank = c(1, 2, 3, 4, 5),
  HISCO_CODE = c(842, 830, 810, 811, 839),
  Occupational_Description = c("Watch, Clock and Precision Instrument Makers",
                               "Blacksmiths, Toolmakers and Machine-Tool Operators",
                               "Woodworkers",
                               "Cabinetmakers",
                               "Blacksmiths, Toolmakers and Machine-Tool Operators")
)

# Create a gt table from the tibble
watchmaker_table <- watchmaker_data %>%
  gt() %>%
  tab_header(
    title = md("**Closest HISCO codes** to Watchmaker Apprentice")
  ) %>%
  cols_label(
    Rank = "Rank",
    HISCO_CODE = "HISCO CODE",
    Occupational_Description = "Occupational Description"
  )

watchmaker_table
```

## Assign HISCOs: Shipyard Assistant Engineer

```{r}

# Create a tibble with the provided data
occupation_data <- tibble(
  Rank = c(1, 2, 3, 4, 5),
  HISCO_CODE = c(43, 24, 982, 42, 959),
  Occupational_Description = c("Ships' Engineers",
                               "Mechanical Engineer",
                               "Ships' Engine-Room Workers",
                               "Ships' Deck Officers and Pilots",
                               "Construction Workers")
)

# Create a gt table from the tibble
occupation_table <- occupation_data %>%
  gt() %>%
  tab_header(
    title = md("**Closest HISCO Codes** to 'Shipyard Assistant Engineer'")
  ) %>%
  cols_label(
    Rank = "Rank",
    HISCO_CODE = "HISCO CODE",
    Occupational_Description = "Occupational Description"
  )

occupation_table

## You can do it too!

![](assets/code.png){fig-align="center" width=100%}

[GitHub repo](https://github.com/j-jayes/who-is-who-in-industry)
